---
title: "Predicting German Horse Race Outcomes using a Benter-Inspired Model"
output: github_document
---

## Introduction

This notebook explores the application of quantitative methods, inspired by the legendary horse racing bettor Bill Benter, to the German horse racing market.

Web-scraped data and a conditional logistic regression model will be leveraged to estimate the probability of each horse winning a race. The aim is to identify potential market inefficiencies and to assess the effectiveness of Benter's approach in the context of German horse racing.

```{r, message=FALSE, warning=FALSE}
library (data.table)
library(survival)
library(tidyverse)
library(zoo)
```


# 1 Horse Racing in Germany

Germany features two primary types of horse racing: Harness racing and flat racing. Steeplechasing and hurdling have largely faded into history. This notebook will concentrate on flat racing.

## 1.1 Betting Market

Betting plays a crucial role in German horse racing, as a portion of the prize money is funded by the parimutuel betting operator's profits.[^parimutuel] Betting in Germany occurs through two primary channels: bookmakers (fixed odds) and the totalizator (parimutuel). For this analysis, the focus lies exclusively on parimutuel odds.

[^parimutuel]: For more information on parimutuel betting, see the [Wikipedia article](https://en.wikipedia.org/wiki/Parimutuel_betting).



## 1.2 Takeout

In parimutuel betting, the track retains a commission known as takeout. In Germany, the takeout for win and place markets is 15%. This analysis will concentrate solely on the win market.


# 2 Bill Benter's Approach

Inspired by Bolton and Chapman's (1986) seminal paper, Bill Benter employed a conditional logistic regression model to predict horse racing outcomes in Hong Kong.^[See Bolton, R.N., & Chapman, R.G.(1986). Searching for  positive returns at the track: A multinomial logistic regression model for handicapping horse races. Management Science, 32(8), pp. 1040-1060.] His innovative approach incorporated the public's estimate, as reflected in betting odds, into his model.^[See Benter, W. (1994). Computer-based horse race handicapping and wagering systems: A report. In: Efficiency of Racetrack Betting Markets, pp. 183-198] Following a similar path, we'll attempt to identify market inefficiencies within German horse racing. 


# 3 Data

This analysis utilizes German horse racing data collected via web scraping from the official website of Deutscher Galopp e.V. The dataset encompasses race results, race details, and betting odds spanning from 2002 to the present. However, due to a significantly higher takeout rate before 2019, only races from 2019 onwards are included in model training and testing. Data prior to 2019 was leveraged for feature engineering, drawing inspiration from the work of Bolton and Chapman (1986). The web scraping process is detailed in the [data_acquisition folder](../data_acquisition/), while the feature engineering steps can be found in the [data_processing folder](../data_processing/).

```{r}
# Import data
races <- readRDS("../data/processed/engineered_features.Rds")
```


## 3.1 Feature Descriptions

### Horse-related features

* **`hosr730`**: Horse's strike rate in the last 2 years
* **`hosr`**: Horse's career strike rate
* **`homean4sprat`**: Horse's mean speed rating in the last 4 races
* **`homeanearn365`**: Horse's mean earnings per race in the last 365 days
* **`holastsprat`**: Horse's last speed rating
* **`hofirstrace`**: Indicator if it's the horse's first race
* **`hodays`**: Number of days since the horse's last race
* **`draweffect_median`**: The median estimated disadvantage (in terms of lengths) associated with the horse's starting stall compared to the innermost stall
* **`gag`**: Horse's handicap rating before the race
* **`gagindicator`**: Takes the value 1 if the horse's current handicap rating is lower than its rating at the time of its most recent win, 0 otherwise
* **`blinkers1sttime`**: Indicator if the horse is wearing blinkers for the first time
* **`weight`**: Weight carried by the horse


### Jockey-related features

* **`josr365`**: Jockey's strike rate in the last 365 days
* **`jowins365`**: Number of wins for the jockey in the last 365 days


### Trainer-related features

* **`trsr`**: Trainer's strike rate


### Other features

* **`odds`**: Betting odds for the horse



```{r}
# Define a vector to store the names of features used in the model
features <- c(
  # Horse-related features
  "hosr730", "hosr", "homean4sprat", "homeanearn365", "holastsprat",
  "hofirstrace", "hodays", "draweffect_median", "gag", "gagindicator", 
  "blinkers1sttime", "weight",
  
  # Jockey-related features
  "josr365", "jowins365", 
  
  # Trainer-related features
  "trsr",
  
  # Other features
  "odds"
)
```



## 3.2 Filtering and Preparing the Data

This section prepares the data for model training and prediction. Jump races and stakes races are excluded as the focus is on flat races and specifically "Ausgleich IV" handicap races. "Ausgleich IV" races are the lowest class of racing in Germany and those races are run very frequently with many observations per horse in a year. Additionally, races before 2019 are removed due to a significantly different takeout rate at that time. The data is further filtered to ensure data integrity by removing races with missing odds, homean4sprat values, or stall numbers. Lastly, races with dead heats are excluded and the data is sorted by date and time.

```{r}
# Finding races with dead heats
dead_heat_races <- races %>% 
  group_by(dg_raceid, position) %>%
  filter(position == 1) %>% 
  summarise(
    position1_count = n()
  ) %>%
  ungroup() %>% 
  filter(position1_count > 1) %>% 
  pull(dg_raceid) 

# Finding races where homean4sprat is missing
homean4sprat_missing_races <- races %>% 
  filter(is.na(homean4sprat)) %>% 
  pull(dg_raceid)

# Finding races where hostall is missing
missing_stall_races <- races %>%  
  filter(is.na(hostall)) %>% 
  pull(dg_raceid)


data <- races %>% 
  # Filter races based on type, class, surface, and date
  filter(
    race_type == "flat",
    race_class_old == "Ausgleich IV",
    surface == "Turf",
    date_time > "2019-01-01 01:00:00"
  ) %>% 
  # Select the necessary columns
  select(
    all_of(
      c(
        features,
        "dg_raceid", "date_time", "win", "dg_horseid", "horse", "hostall"
      )
    )
  ) %>%
  # Handle missing values and ensure data integrity
  filter(
    !dg_raceid %in% dead_heat_races,
    !dg_raceid %in% homean4sprat_missing_races,
    !dg_raceid %in% missing_stall_races,
    !is.na(odds)
  ) %>% 
  mutate(
    hodays = ifelse(hofirstrace == 1, 0, hodays)
  ) %>% 
  arrange(date_time) %>% 
  group_by(dg_horseid) %>% 
  # For each horse use last measured speed rating if last speedrating is missing
  mutate(holastsprat = na.locf(holastsprat, na.rm = FALSE)) %>% 
  ungroup()
```



# 4 Training and Testing the Model

The dataset is split into a training set and a test set to evaluate the model's performance on unseen data. The training set includes races before January 1, 2021, while the test set contains races from that date onwards. This split allows us to assess how well the model generalizes to new races and avoids overfitting to the training data.

```{r}
# Create training dataset using races before 2021
train_data <- data %>% 
  filter(date_time < "2021-01-01 01:00:00")

# Create test dataset using races after 2021
test_data <- data %>% 
  filter(date_time > "2021-01-01 01:00:00") %>% 
  data.table()
```


## 4.1 Training the Model

A conditional logistic regression model (`clogit`) is employed to predict the probability of a horse winning a race. This model is wellsuited for analyzing data with multiple observations within groups (horses within races) and allows us to account for the inherent  dependencies within each race. This approach is consistent with Bill Benter's strategy in horse race prediction. But in contrast to Benter's two-step approach to incorporate the public estimate (the odds), here a one-step approach is utilized. 



```{r}
# Construct the model formula using the selected features
model_formula <- as.formula(
  paste(
    "win",
    paste(paste(features, collapse = " + "), "strata(dg_raceid)", sep = " + "),
    sep = " ~ "
  )
)
print(model_formula)
```

```{r}
# Fit the conditional logistic regression model
model <- clogit(
  model_formula,
  data = train_data, method = "exact"
)

# Print model summary
summary(model)
```

The estimated coefficients are extracted from the model summary for use in subsequent predictions on the test data.

```{r}
# Extract coefficients from the model summary
coeffs <- as.vector(summary(model)$coefficients[, 1])
coeffs
```



## 4.2 Testing the Model

The performance of the trained model on unseen data will be evaluated by generating predictions for races in the test set (races after January 1, 2021). Based on these predictions expected values are calculated. A straightforward betting strategy is to bet on the horse with the highest positive expected value in a race. 


Predictions

```{r}
predictions <- test_data[, prediction := as.matrix(test_data[, ..features]) %*% coeffs]
predictions <- predictions %>% 
  group_by(dg_raceid) %>% 
  mutate(
    exp_prediction = exp(prediction),
    sum_exp_prediction = sum(exp_prediction),
    my_prob = exp_prediction / sum_exp_prediction,
    expected_value = my_prob * (odds - 1) - (1 - my_prob)
  ) %>% 
  filter(
    expected_value > 0,
    expected_value == max(expected_value)
  )  %>% 
  ungroup() %>% 
  mutate(
    earnings = ifelse(
      win == 1, odds - 1, -1
    )
  )

sum(predictions$earnings)
```

```{r}

```




