---
title: "Using Bill Benter's Methods in German Horse Racing"
output: github_document
---

```{r, message=FALSE, warning=FALSE}
library (data.table)
library(survival)
library(tidyverse)
library(zoo)
```


# 1 Horse Racing in Germany

## 1.1 Betting Market

Germany features two primary types of horse racing: Harness racing and flat racing. Steeplechasing and hurdling have largely faded into history. This notebook will concentrate on flat racing. 

Betting plays a crucial role in German horse racing, as a portion of the prize money is funded by the parimutuel betting operator's profits. Betting in Germany occurs through two primary channels: bookmakers (fixed odds) and the totalizator (parimutuel). For this analysis, the focus lies exclusively on parimutuel odds.


## 1.2 Takeout

In parimutuel betting, the track retains a commission known as takeout. In Germany, the takeout for win and place markets is 15%. Our analysis will concentrate solely on the win market.


# 2 Bill Benter's Approach

Inspired by Bolton and Chapman's (1986) paper, Bill Benter employed a conditional logistic regression model to predict horse racing outcomes in Hong Kong. His innovative approach incorporated the public's estimate, as reflected in betting odds, into his model. Following a similar path, we'll attempt to identify market inefficiencies within German horse racing. 


# 3 Data

The data used in this analysis has been acquired by web scraping. German horse racing results since 2002 up until 2024 have been gathered. But only the results since 2019 will be used in training and testing the model because before 2019 the takeout rate was much higher than 15%. Data before 2019 has however been used to construct the necessary features. Similar features to those mentioned by Bolton and Chapman (1986) have been engineered.

```{r}
# Import data
races <- readRDS("../data/processed/engineered_features.Rds")
```


## 3.1 Feature Descriptions

### Horse-related features

* **`hosr730`**: Horse's strike rate in the last 2 years
* **`hosr`**: Horse's career strike rate
* **`homean4sprat`**: Horse's mean speed rating in the last 4 races
* **`homeanearn365`**: Horse's mean earnings per race in the last 365 days
* **`holastsprat`**: Horse's last speed rating
* **`hofirstrace`**: Indicator if it's the horse's first race
* **`hodays`**: Number of days since the horse's last race
* **`draweffect_median`**: The median estimated disadvantage (in terms of lengths) associated with the horse's starting stall compared to the innermost stall
* **`gag`**: Horse's handicap rating before the race
* **`gagindicator`**: Takes the value 1 if the horse's current handicap rating is lower than its rating at the time of its most recent win, 0 otherwise
* **`blinkers1sttime`**: Indicator if the horse is wearing blinkers for the first time
* **`weight`**: Weight carried by the horse


### Jockey-related features

* **`josr365`**: Jockey's strike rate in the last 365 days
* **`jowins365`**: Number of wins for the jockey in the last 365 days


### Trainer-related features

* **`trsr`**: Trainer's strike rate


### Other features

* **`odds`**: Betting odds for the horse



```{r}
# Define a vector to store the names of features used in the model
features <- c(
  # Horse-related features
  "hosr730", "hosr", "homean4sprat", "homeanearn365", "holastsprat",
  "hofirstrace", "hodays", "draweffect_median", "gag", "gagindicator", 
  "blinkers1sttime", "weight",
  
  # Jockey-related features
  "josr365", "jowins365", 
  
  # Trainer-related features
  "trsr",
  
  # Other features
  "odds"
)
```



## 3.2 Filtering and Preparing the Data

This section prepares the data for model training and prediction. Jump races and stakes races are excluded as the focus is on flat races and specifically "Ausgleich IV" handicap races. "Ausgleich IV" races are the lowest class of racing in Germany and those races are run very frequently with many observations per horse in a year. Additionally, races before 2019 are removed due to a significantly different takeout rate at that time. The data is further filtered to ensure data integrity by removing races with missing odds, homean4sprat values, or stall numbers. Lastly, races with dead heats are excluded and the data is sorted by date and time.

```{r}
# Finding races with dead heats
dead_heat_races <- races %>% 
  group_by(dg_raceid, position) %>%
  filter(position == 1) %>% 
  summarise(
    position1_count = n()
  ) %>%
  ungroup() %>% 
  filter(position1_count > 1) %>% 
  pull(dg_raceid) 

# Finding races where homean4sprat is missing
homean4sprat_missing_races <- races %>% 
  filter(is.na(homean4sprat)) %>% 
  pull(dg_raceid)

# Finding races where hostall is missing
missing_stall_races <- races %>%  
  filter(is.na(hostall)) %>% 
  pull(dg_raceid)


data <- races %>% 
  # Filter races based on type, class, surface, and date
  filter(
    race_type == "flat",
    race_class_old == "Ausgleich IV",
    surface == "Turf",
    date_time > "2019-01-01 01:00:00"
  ) %>% 
  # Select the necessary columns
  select(
    all_of(
      c(
        features,
        "dg_raceid", "date_time", "win", "dg_horseid", "horse", "hostall"
      )
    )
  ) %>%
  # Handle missing values and ensure data integrity
  filter(
    !dg_raceid %in% dead_heat_races,
    !dg_raceid %in% homean4sprat_missing_races,
    !dg_raceid %in% missing_stall_races,
    !is.na(odds)
  ) %>% 
  mutate(
    hodays = ifelse(hofirstrace == 1, 0, hodays)
  ) %>% 
  arrange(date_time) %>% 
  group_by(dg_horseid) %>% 
  # For each horse use last measured speed rating if last speedrating is missing
  mutate(holastsprat = na.locf(holastsprat, na.rm = FALSE)) %>% 
  ungroup()
```



# Train Data

```{r}
train_data <- data %>% 
  filter(date_time < "2021-01-01 01:00:00")
```

Model

 + hosr + homean4sprat + homeanearn365 + holastsprat + josr365 + 
    jowins365 + weight + hostall + hono + hofirstrace

```{r}

model_formula <- as.formula(
  paste(
    "win",
    paste(paste(features, collapse = " + "), "strata(dg_raceid)", sep = " + "),
    sep = " ~ "
  )
)
print(model_formula)
```



```{r}
model <- clogit(
  model_formula,
  data = train_data, method = "exact"
)
summary(model)
```


```{r}
coeffs <- as.vector(summary(model)$coefficients[, 1])
coeffs
```



# Test Data

```{r}
test_data <- data %>% 
  filter(
    date_time > "2021-01-01 01:00:00" 
  ) %>% 
  data.table()
```

Predictions

```{r}
predictions <- test_data[, prediction := as.matrix(test_data[, ..features]) %*% coeffs]
predictions <- predictions %>% 
  group_by(dg_raceid) %>% 
  mutate(
    exp_prediction = exp(prediction),
    sum_exp_prediction = sum(exp_prediction),
    my_prob = exp_prediction / sum_exp_prediction,
    expected_value = my_prob * (odds - 1) - (1 - my_prob)
  ) %>% 
  filter(
    expected_value > 0,
    expected_value == max(expected_value)
  )  %>% 
  ungroup() %>% 
  mutate(
    earnings = ifelse(
      win == 1, odds - 1, -1
    )
  )

sum(predictions$earnings)
```

```{r}

```




